---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r init, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)

# Set the seed of the random generator so that the output
# of random data does not always change
set.seed(12345)
```

# blockrand2

This R package helps you to create block randomised treatment lists for clinical studies.It is highly inspired by and quite similar to the existing package `blockrand` that is available on [CRAN](https://cran.r-project.org/web/packages/blockrand/index.html), but was developed independently. Compared to `blockrand` it offers additional functionality:

- generation of test data,
- stratification of patient's data,
- generation of a report (html, pdf and/or docx).

For the output created by the following example session see here: [HTML](https://github.com/hsonne/blockrand2/tree/master/inst/extdata/randomList.html)
[PDF](https://github.com/hsonne/blockrand2/tree/master/inst/extdata/randomList.pdf)
[DOCX](https://github.com/hsonne/blockrand2/tree/master/inst/extdata/randomList.docx)

## Installation

You can install the latest development version of the package `blockrand2` from 
github with

```R
# install.packages("devtools")
devtools::install_github("hsonne/blockrand2")
```

## Usage

First of all, you need to load the package:

```{r load_package}
library(blockrand2)
```

### Define Document Header

Start by defining the title of the study, the author's name and the date in form
of a list with list elements `title`, `author` and `date`. These information 
will appear on top of the created document(s).

```{r define_header}
# Define the title of the study, the author and the date
header <- list(
  title = "Randomisation lists for the study: *Jokes against craziness*",
  author = "Hauke Sonnenberg",
  date = "2016-08-27"
)
```

### Define Stratum Variables and Treatments

Next, define the names and possible values of the variables that define 
different strata into which to group the patients that take part in your study:

```{r define_variables}
# Define stratum variables
strataVars <- list(
  sex = c("male", "female"),
  crazyness = c("weak", "medium", "strong")
)
```

Define the different treatments that are to be applied to the patients with a
short name and a title in the form of a named character vector:

```{r define_treatments}
# Define the treatments
treatments <- c(
  joke = "Tell funny jokes",
  nojoke = "Keep serious!"
)
```

### Provide Patient's Data

Of course, we need information on our patients to which we want to assign the
different treatments. Let's create some random testdata based on the stratum
variables already defined. Therefore, the package provides the function 
`createTestdata()` to which you pass the definition of stratum variables as well
as the number of records to be created:

```{r provide_testdata}
# Create some patient's testdata using the stratum variables already defined
patients <- createTestdata(strataVars, 40)
```

Let's have a look at the first lines of the created table:

```{r show_testdata}
head(patients)
```

Once you have prepared a file (e.g. in CSV format) containing the real data you 
will replace the above line with something like this:

```{r provide_real_data, eval = FALSE}
# Read patient's data from a CSV file
patients <- read.csv("patients.csv")
```

Make sure that the number of columns and the column names in your file are the 
same as are provided by `createTestdata()` 
(here: `r paste0("*", names(patients), "*", collapse = ", ")`). 
Otherwise the following steps will not work properly.

### Create the Randomisation List Document(s)

That's more or less it. All you have to do now is to call the function 
`createRandomisationDoc()` to which you have to pass the `patient`'s data, the
definition of `strataVars` and the `treatments` as well as the information going
into the `header` of the document(s) to be created. The argument `newpage` 
controls (only for output format "pdf") whether a new page is to be started 
after each list showing the patient's data and their assigned treatments for one 
stratum.

```{r create_docs_hide, echo=FALSE, message=FALSE, warning=TRUE}
trash <- capture.output(files <- createRandomisationDoc(
  patients = patients,
  strataVars = strataVars,
  treatments = treatments,
  header = header,
  newpage = TRUE
))
```

```{r create_docs_show, eval = FALSE}
# Create the randomisation list documents
files <- createRandomisationDoc(
  patients = patients,
  strataVars = strataVars,
  treatments = treatments,
  header = header,
  newpage = TRUE
)
```

The full paths to the generated files are returned invisibly. Since we stored
these paths in the variable `files` (being a `list`) we can review these paths

```{r show_files}
# Show the paths of the created files
files
```

```{r copy_files, echo = FALSE, include = FALSE}
# Copy html file to inst/extdata
for (file in files) {
  file.copy(file, "inst/extdata")
}
```

and use them directly to open the created files from within `R` with their appropriate applications (if available):

```{r open_files, eval = FALSE}
# Open the html file in the default browser
browseURL(files$html)

# Open the pdf file in the default PDF viewer
system(paste(getOption("pdfviewer"), files$pdf))
```

By default, the randomisation list document is created in three formats: 
`"html", "word"` and `"pdf"`. This is controlled by the formal argument `format`
of `createRandomisationDoc()`. If, for example you do not have LaTeX installed
which is the requirement to create the document in PDF format, you may select to
e.g. only generate `"word"` format:

```{r create_docs_word, eval = FALSE}
# Create the randomisation list documents
createRandomisationDoc(
  patients = patients,
  strataVars = strataVars,
  treatments = treatments,
  header = header,
  format = "word"
)
```
